MONDES/WORLDS documentary connaissance du monde

28000 light year from the center of the Milky Way, on the Orion, a nebula collap
se on itselft.  A chain of nuclear reaction takes place, the temperature reach 15million degree, a star is born, the Sun. The story of our solar system has begin.

15M yr after started being formed, the sun was still as protostar surrounded by a vast gas of smoked matter.
After 10M yr, 4 giant planets took shape (Uranus, Neptune, Saturne, Jupiter), gas giants formed rapidly before the gas from rthe nebula evaporated reapidly forever. Jupiter was 50 times bigger than earth, thre solar sustem was mutating. Having migrated toward the Sun, Jupyter and Saturn then pull away from the star and move toward the exterior, disturbing most the asteroid belt. Jupiter and Saturn orbits expanded, deviating those of Uranus and Neptune. The latter between Jupiter and Neptunr was pushed out to the edge of the solar system. 

Clsoed to thre Sun, lighter and more volatided was replaced by heavier matter: rock dust and matter. These grains ofg fdust collide, fuse, become peebles. P{eebels then became rocks. Rockes become heavier enough to create their own gravity and attract matter, giving births to telluric planets. There are 4 of these: Mars, Earth, Venus, Mecury

There may be there are more telluric planets when the Sun was still young. Perhaps, they are pushed out to the edge of the sollar system or swalloed by the Sun. 

Our planet was formed around 13M yr, 

then when the earh is 50Myr old, our planet is under a catalysm, collided with another telluric planet. Debris formed in an orbit, accumulate and form the moon. At this time, the earth looks so inhabitable, but life has appeared: 4 billion yrs ago,  

Why life has appear on earth?
First, Earth is at the hear of the solar system, at the habitablr zone. temperature is good for water to be in gaseous, liquid, solid.
And from wter, lfie might have formed. 
Water cover 75\% planet's surface. Moon provides some clue. Moon rok from the Apollo misison. Moon was bombarded intensely between 4.1 and 3.9 billion yr ago due to the meteripot and comets showers. Earth also endured the same.

Meteroite and Comets contains entirely ice bring water to Earth. It added to 30\% of water that was orignailed when the earth was formed. Whic form first pond, lake, sea, ocean. Gradulaty, it stays anfd the earth cool downs.

Then, 3.8 billion yr ago, first single cell organism apperas in hot volcanic spring. Hydrothermal spring with the ideal places for succession phases of dry and humid that help contration and building molecules.  Then, thansk to photosynthesis by Sun, ther atmoshere becomes rich in oxygen. allow microrganism develoepd in the ocean. From then on, life becoems remarkable evolution.,

Could ligfe form elsewhere?

Mecuri is the smallest planet and closet to the sun, very similAR in size to the moon. Condition is extreme, Temperteu reach K on the side facing the sun, -150K on the dark side. Mercury has no atmosphere, therefore preyed to meterotie and asteroid, posted marks on the landscapes.
Densese planet: 2018, bepbi colombo orbits around mervcury to understand how planets are form. Two saatellive were put to study the magnetic field, to udnerstand relluric evulition and proimity to the Sun. Mercuri is not a death planet, geologically activem  despite its look. Mercuri is sinking because its kernel is cooling down. In 4 billion year, it losst 7km of radius. 

Bepbi cross path of Venus, 2nd planet in the solar system. same size as Earth, may once has life. dream of habiltable planets. 1970, this image of Venus, verenus unvreal strcuturer of the atmoshere and characteristics of  Venus. they discovered a world of hellish conditions. Temperature og 460 C, 95\% level of CO2, thick cloud ofd sulfur gas, no magnetic field to procte from cosmic ray radiaion and solar wind. Magenalc probe reveals the surface by radar. The surface is vyoung and volcanic, vast laval plane and thoudns of volcanic peaks. High peask is 8000 m high. 2020, detect phosgine in venus atmospheere, colorless odouress highly toxic gas, produced by microorganism on Earth. Unexpededc in Venus clouds because it can be destroyed in the atmosthere. They postulate that it is form by a life form in the Venusian cloud. A new mission of NASA, ballon 50km from the atmosphere,75 C degree, build a city of ballons to measure in siute the condition of venus clpouds.

Mars: 4th planet. A martian year is 2 as long as on earth but the day is the sam as earth. In ancitent time, Mars is red therefore has the name of the Greek god of war. Red corlour is deu to high level of ....., nother word, mars is rusting. Billion of years agom Mars was also colldide with an embryonic planet about 1/3 of itself. Demois (11km accross ) and Phobos (6 km accross, 300 time smaller than moon) were formed during the catalys: smallest natural satellite in the solar system. Phobos move towards mars by 1.8km/century, will get hotter, fall apart and form debris ring around Mars, similar to SAturn. Demos moves away from Mars, getting less attract from mars and drift to the space. Mars can lsot both their childerns.

1948, Gerard Kuiper discover ice in mars polar cap. Presence of water in ther form of river bed by images... Orignall, mars was hotter and covered by volcabos, and especialle liqyud water. Water flow acros the planet for 2-3.5 billion years. Mars might have great lakes, and even ocean.  Another time. planet become colder, iuts metallic hard froze, its magnetic shield collapse alllowng solar wind to dry out its atmoshere and caus its water to evaporate. Could there were some form of life ? 
NASA: perseverence misison, a tiny helicoper, Tianwen I and rover, Exomars w2020 rosalind franklin. search for organix matter in the subsoild and make first contact with any life form.

Organic matter are found from asteroid belt bettwen mars and juputer: stones, ricks stardust, planets never forms, eventually collide. rocks change name to meterotie when hit the EArth.2001, 600000 bodies has been recorded. smallest is 2m in diameter. The biggest Veesta 580 km. Ceres has been  long classirf: Dwarf planet. Ceres gravity is not signgicant enoug to clear the bodies arounds.

Asteroid mission: Lucie, 

these asterod never aggloberate imto planets due to Jupiter influence, the lartgest plaetn in our solar ysysten. It is made of uniquely of gas. a probr can go through the plamnet and go to the other way. Jupiyer is a failed star, not enough matter in the primitive nevuae to achive critical mass to cause nuclear fusion. Cassini probe studied planet, Juno probe studi Jupyer origin. Jupityer retian original compositions. Jupiter has over 100 satellite. the 4 biggest: Ganymede, Callisto, Io, Europe. Thesr moon show similarity with the Eartth. Europa is the most intriguring moon. Hubble discovered water in steam state: gaseous. Europa has also a vast ocean with liquid water beneath. Clipper 2025 fly aorund Europa to analyze composion of water vapour and find indirector tracer of life. 

Life coudl exit on the Saturn biggtest Moon: titan. This Moon has an atmosphere. This moor is slightly bigger than Mecuri. only solid body system as a dense atmpsphere,  contains 95\% of Nitrooxide, climate similarto our earth, the only one as liquid methan on its surface. Aeroal exploration mission : dragonfly mission will expore titan equator.  titan is a laboraty of organic molecluar factory. Along with Earth, it is the only body that has liquid expended on the surface. Titan hasws 10 times more liquid expense than Earth ocean.  50 km benith its surface is a global lake 200km deep.  NASA create spatial submarine: 6m lomngg. 
Enceledous is a small satellite. Its surface is very birhg, reflect almsot all light. water vaour around south pole, also has liqud water beneath its sirfuace, shooting water to groter,adn thown into space . life could exists on this moon. 

Uranis has 13 rings, and Neptunes are 2 ice giants, mainly compsoed of liquid elements. Nrptune mass is signigicatnt greateer Uranus tilled arose as if it is rolling around the Sun. Trian neptune biggest satellite, obit in the opposite direction of its plaent. 

Neptune, pluton moon

Eris heavier than pluto let pluto 

ceris pluto, eris: dwarf planets

pluto composiiotn is unkownn: but simulatio show an ocena, charon is the moon: big satellite half of pluto.

kuiprt belt, oort clooud

only vouyager 1, 2 has come that far

Exoplanet 51 pegasi B

kepler 186F

omnly 24 exoplanets migjh be hbitavle KOI5715


 















